---
title: "NHS 2020: Forecasting multiple seasonality"
subtitle: "Multiple seasonal forecasting with fable"
author: "Mitchell O'Hara-Wild"
date: '25/11/2020'
output:
  xaringan::moon_reader:
    chakra: ./libs/remark-latest.min.js
    css: ["././libs/slides.css", "././libs/animate.css"]
    lib_dir: ./libs
    seal: false
    anchor_sections: false
    nature:
      highlightStyle: github
      ratio: '16:9'
      highlightLines: true
      countIncrementalSlides: false 
      beforeInit: ["./libs/jquery/jquery.min.js", "./libs/slides.js"]
---
class: inverse
background-image: linear-gradient(to right, rgba(150, 150, 150, .1), rgba(150, 150, 150, .4)), url("resources/hourglass.jpg")
background-size: cover

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE, width = 120)

library(tidyverse)
library(knitr)
library(kableExtra)
library(fontawesome)
library(lubridate)
library(htmltools)

library(tsibble)
library(fasster)
library(fable)

opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE, comment = "#>",
  fig.path = 'figure/', cache.path = 'cache/', fig.align = 'center',
  fig.width = 12, fig.height = 4, fig.show = 'hold',
  cache = TRUE, external = TRUE, dev = 'svglite', dev.args = list(bg = "transparent")
)

mp4_vid <- function(src){
  HTML(
    paste0(
      '<video autoplay>
        <source src="', src, '" type="video/mp4">
      </video>'
    )
  )
}

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

theme_set(
  theme_grey(base_size = 16) +
  theme(
    legend.position = "bottom",
    plot.background = element_rect(fill = "transparent"),
    legend.background = element_rect(fill = "transparent")
  )
)
```


```{css, echo=FALSE}
/* custom.css */
.left-code {
  color: #777;
  width: 40%;
  height: 92%;
  float: left;
}
.right-plot {
  width: 58%;
  float: right;
  padding-left: 1%;
}
```


.title[fable]
.sticker-float[![fable](resources/fable.svg)]

## Forecasting with multiple seasonality

.bottom[
### Mitchell O'Hara-Wild (`r fa("twitter", fill="#1da1f2")`[@mitchoharawild](https://twitter.com/mitchoharawild))
### 25 November 2020
### Slides @ [slides.mitchelloharawild.com/nhs2020](https://slides.mitchelloharawild.com/nhs2020)
]

---
class: inverse
background-image: linear-gradient(to right, rgba(150, 150, 150, .1), rgba(150, 150, 150, .4)), url("resources/timezone.jpg")
background-size: cover

.title[Hello!]

---
class: center
background-image: linear-gradient(to right, rgba(255, 255, 255, .1), rgba(150, 150, 150, .1)), url("resources/season.jpg")
background-size: cover

## Understanding multiple seasonality

---
background-image: linear-gradient(to right, rgba(255, 255, 255, .1), rgba(150, 150, 150, .1)), url("resources/season.jpg")
background-size: cover

.center[
## What is seasonality?
]

--

> A consistent pattern which repeats over a fixed period of time.

--

.box-12.opaque[
```{r ped-seasonality}
pedestrian %>% 
  filter(Sensor == "Southern Cross Station",
         between(Date, parse_date("2015-04-13"), parse_date("2015-04-17"))) %>% 
  autoplot(Count) + 
  scale_x_datetime(date_labels = "%A", date_breaks = "1 day") +
  labs(y = "Number of pedestrians", x = "Time", title = "Pedestrian traffic at Southern Cross Station (Victoria, Australia)")
```
]


---
background-image: linear-gradient(to right, rgba(255, 255, 255, .1), rgba(150, 150, 150, .1)), url("resources/season.jpg")
background-size: cover

.center[
## What is *multiple* seasonality?
]

--

> Two or more seasonal patterns.

--

.box-12.opaque[
```{r ped-multi-seasonality}
pedestrian %>% 
  filter(Sensor == "Southern Cross Station",
         yearmonth(Date) == yearmonth("2015 Apr")) %>% 
  autoplot(Count) + 
  scale_x_datetime(date_labels = "%d %b %Y", date_breaks = "5 days") +
  labs(y = "Number of pedestrians", x = "Time", title = "Pedestrian traffic at Southern Cross Station (Victoria, Australia)")
```
]

---
class: inverse
background-image: linear-gradient(to right, rgba(50, 50, 50, .5), rgba(50, 50, 50, .4)), url("resources/tree.jpg")
background-size: cover

.center[
## How can seasonal patterns be modelled?
]

<hr>

--

<br><br>
.center[
### Seasonal dummy variables

### Lagged terms

### Fourier terms

### Exogenous regressors
]

---

## Modelling seasonal patterns with...

## Seasonal dummy variables

.pull-left[
Produces distinct intercepts for each discrete season.

✅ Simplest to interpret and produce.

❌ Requires many terms for longer periods.

❌ Poorly handles complex seasonality.
]

.pull-right[
```{r dowdummy, echo=FALSE}
df <- matrix("0", nrow=13, ncol=6)
df[1:6,] <- paste(diag(6))
df[8:12,] <- paste(diag(6)[1:5,])
colnames(df) <- paste("$$d_{",1:6,",t}$$",sep="")
df <- as_tibble(df)
df$day <- rep(c("Monday","Tuesday","Wednesday","Thursday","Friday","Saturday","Sunday"),2)[1:13]

dfrows <- 9
df <- df[seq(dfrows),]
df[dfrows,] <- "..."
gt::gt(df, rowname_col = "day")
```
]

---

## Modelling seasonal patterns with...

## Lagged terms

.pull-left[
*'Today will be similar to yesterday.'*

✅ Very difficult to beat accuracy of.

❌ Limited forecast horizon.

❌ Cannot handle non-integer seasonality.
]

.pull-right[
```{r lag-plot, fig.height = 10}
pedestrian %>% 
  filter(Sensor == "Southern Cross Station",
         between(Date, parse_date("2015-04-13"), parse_date("2015-04-17"))) %>% 
  autoplot(Count) + 
  geom_line(aes(y = lag(Count, 24)), alpha = 0.5, colour = "red") + 
  scale_x_datetime(date_labels = "%A", date_breaks = "1 day") +
  labs(y = "Number of pedestrians", x = "Time", title = "Pedestrian traffic at Southern Cross Station (Victoria, Australia)")
```
]

---

## Modelling seasonal patterns with...

## Fourier terms

$$x_{1,t} = \sin\left(\textstyle\frac{2\pi t}{m}\right), x_{2,t} = \cos\left(\textstyle\frac{2\pi t}{m}\right),$$

$$x_{3,t} = \sin\left(\textstyle\frac{4\pi t}{m}\right), x_{4,t} = \cos\left(\textstyle\frac{4\pi t}{m}\right), \ldots$$
```{r fourier-plot, fig.height=3}
ped_fourier <- tsibble::pedestrian %>% 
  slice(24:(24*8) - 12)
bind_cols(
  ped_fourier,
  fable:::fourier.tbl_ts(ped_fourier, 24*7, 2)
) %>% 
  pivot_longer(ends_with("168"), names_to = "term", values_to = "fourier") %>% 
  ggplot(aes(x = Date_Time, y = fourier, colour = term)) + 
  geom_line() + 
  labs(x = NULL, y = NULL, colour = NULL)
```

---

## Modelling seasonal patterns with...

## Fourier terms

.pull-left[
Captures seasonal distribution with a continuous periodic functions.

✅ Handles non-integer seasonality (days in year, weeks in month).

✅ Flexibility of seasonal pattern controllable by number of harmonics.

❌ More complicated to interpret.
]

.pull-right[
```{r fourier-plot-2, ref.label="fourier-plot", fig.height = 10}
```
]

---

## Modelling seasonal patterns with...

## Exogenous regressors

.pull-left[
Capture seasonality by providing underlying factors like temperature, and events.

✅ Better handling of non-seasonal anomalies.

❌ Data may be hard to find.

❌ Future values must be forecasted.
]

.pull-right[
```{r exog-plot, fig.height=10}
pedestrian %>% 
  filter(Sensor == "Southern Cross Station",
         yearmonth(Date) == yearmonth("2015 Apr")) %>% 
  group_by(Date) %>% 
  mutate(workday = max(Count) > 2000) %>% 
  ungroup() %>% 
  autoplot(Count) + 
  geom_rect(aes(xmin = lag(Date_Time), xmax = Date_Time, ymin = -Inf, ymax = Inf, fill = workday), alpha = 0.3) + 
  geom_line() + 
  scale_x_datetime(date_labels = "%d %b %Y", date_breaks = "5 days") +
  labs(y = "Number of pedestrians", x = "Time", title = "Pedestrian traffic at Southern Cross Station (Victoria, Australia)")
```
]
---
class: inverse
background-image: linear-gradient(to right, rgba(50, 50, 50, .5), rgba(50, 50, 50, .4)), url("resources/tree.jpg")
background-size: cover

.center[
## Which seasonal method should I use?
]

<hr>

--

<br>

.center[

### For *multiple* seasonality, *multiple* methods may be needed.
<br>

### Use methods that are available and work best for your data!
]

---

## Models for forecasting multiple seasonalities

Capability for including seasonal terms.

<table class="table">
  <thead>
    <tr>
      <th>Model</th>
      <th>Dummies</th>
      <th>Lags</th>
      <th>Fourier</th>
      <th>Regressors</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <td>MLR</td>
      <td class="success">Yes</td>
      <td class="success">Yes</td>
      <td class="success">Yes</td>
      <td class="success">Yes</td>
    </tr>
    <tr>
      <td>GAM</td>
      <td class="success">Yes</td>
      <td class="success">Yes</td>
      <td class="success">Yes</td>
      <td class="success">Yes</td>
    </tr>
    <tr>
      <td>Prophet</td>
      <td class="danger">No</td>
      <td class="success">Yes</td>
      <td class="success">Yes</td>
      <td class="success">Yes</td>
    </tr>
    <tr>
      <td>DSHW</td>
      <td class="success">Yes</td>
      <td class="danger">No</td>
      <td class="danger">No</td>
      <td class="danger">No</td>
    </tr>
    <tr>
      <td>BATS/TBATS</td>
      <td class="success">Yes</td>
      <td class="warning">Partial</td>
      <td class="success">Yes</td>
      <td class="danger">No</td>
    </tr>
    <tr>
      <td>FASSTER</td>
      <td class="success">Yes</td>
      <td class="success">Yes</td>
      <td class="success">Yes</td>
      <td class="success">Yes</td>
    </tr>
    <tr>
      <td>STL + ???</td>
      <td class="warning">???</td>
      <td class="warning">???</td>
      <td class="warning">???</td>
      <td class="warning">???</td>
    </tr>
  </tbody>
</table>

.footnote[(This *non-exhaustive* list includes the most common statistical methods only.)]

---

## Models for forecasting multiple seasonalities


<br>
<table class="table">
  <thead>
    <tr>
      <th>Model</th>
      <th>Flexible</th>
      <th>Speed</th>
      <th>Accuracy</th>
      <th>Decompose Components</th>
      <th>Evolving Terms</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <td>Regression<br>(MLR, GAM, Prophet)</td>
      <td></td>
      <td></td>
      <td></td>
      <td></td>
      <td></td>
    </tr>
    <tr>
      <td>State&nbspSpace<br>(DSHW, TBATS, FASSTER)</td>
      <td></td>
      <td></td>
      <td></td>
      <td></td>
      <td></td>
    </tr>
    <tr>
      <td>Decomposition<br>(STL + ???)</td>
      <td></td>
      <td>&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;</td>
      <td></td>
      <td></td>
      <td></td>
    </tr>
  </tbody>
</table>

---

## Models for forecasting multiple seasonalities


<br>
<table class="table">
  <thead>
    <tr>
      <th>Model</th>
      <th>Flexible</th>
      <th>Speed</th>
      <th>Accuracy</th>
      <th>Decompose Components</th>
      <th>Evolving Terms</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <td>Regression<br>(MLR, GAM, Prophet)</td>
      <td class="warning">Depends</td>
      <td class="warning">Depends</td>
      <td class="warning">Depends</td>
      <td class="warning">Depends</td>
      <td class="warning">Depends</td>
    </tr>
    <tr>
      <td>State&nbspSpace<br>(DSHW, TBATS, FASSTER)</td>
      <td class="warning">Depends</td>
      <td class="warning">Depends</td>
      <td class="warning">Depends</td>
      <td class="warning">Depends</td>
      <td class="warning">Depends</td>
    </tr>
    <tr>
      <td>Decomposition<br>(STL + ???)</td>
      <td class="warning">Depends</td>
      <td class="warning">Depends&nbsp;&nbsp;</td>
      <td class="warning">Depends</td>
      <td class="warning">Depends</td>
      <td class="warning">Depends</td>
    </tr>
  </tbody>
</table>

---

## Models for forecasting multiple seasonalities


<br>
<table class="table">
  <thead>
    <tr>
      <th>Model</th>
      <th>Flexible</th>
      <th>Speed</th>
      <th>Accuracy</th>
      <th>Decompose Components</th>
      <th>Evolving Terms</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <td>Regression<br>(MLR, GAM, Prophet)</td>
      <td class="success">Yes</td>
      <td class="success">Fast</td>
      <td class="warning">Okay</td>
      <td class="warning">Depends</td>
      <td class="danger">No</td>
    </tr>
    <tr>
      <td>State&nbspSpace<br>(DSHW, TBATS, FASSTER)</td>
      <td class="danger">No</td>
      <td class="danger">Slow</td>
      <td class="warning">Okay</td>
      <td class="success">Yes</td>
      <td class="success">Yes</td>
    </tr>
    <tr>
      <td>Decomposition<br>(STL + ???)</td>
      <td class="warning">Depends</td>
      <td class="warning">Moderate</td>
      <td class="warning">Okay</td>
      <td class="success">Yes</td>
      <td class="warning">Depends</td>
    </tr>
  </tbody>
</table>

.footnote[(This table is a *massive* generalisation, everything depends on data and model complexity.)]

---
class: center

## Forecasting multiple seasonality in R

.sticker[![fable](resources/fable.svg)]

--

.animated.fadeIn[
.sticker[![tsibble](resources/tsibble.svg)]
.sticker[![tsibbledata](resources/tsibbledata.svg)]
.sticker[![feasts](resources/feasts.svg)]

## [tidyverts.org](http://www.tidyverts.org)
]

---
class: top

```{r}
htmltools::tags$iframe(src = "https://otexts.com/fpp3/a-tidy-forecasting-workflow.html", 
                       width = "100%", height = "640px")
```

---
class: inverse, center
background-image: linear-gradient(to right, rgba(50, 50, 50, .5), rgba(50, 50, 50, .5)), url("resources/disk.jpg")
background-size: cover

.title[Tidy]

<hr>
<br>

## Data preparation

---
class: top

# Hospital admissions

```{r, echo = TRUE}
fs::dir_tree("~/github/webinar-msc-nhs/")
```

.footnote[*https://github.com/bahmanrostamitabar/webinar-msc-nhs/*]

---
class: top

# Admissions event data

```{r admission-raw, echo = TRUE}
admissions <- read_csv(
  file = "~/github/webinar-msc-nhs/data/ae_uk.csv",
  col_types = cols(arrival_time=col_datetime(format = "%d/%m/%Y %H:%M"))
) %>%
  select(-ID) # ID is a redundant row identifier 
print(admissions)
```

---
class: top

# Holiday data

```{r holiday-raw, echo = TRUE}
holidays <- read_csv(
  file = "~/github/webinar-msc-nhs/data/holiday.csv",
  col_types = cols(date=col_date(format = "%d/%m/%Y"))
) %>% 
  filter(!is.na(date)) # Some extra empty rows exist that need removing
print(holidays)
```

---
class: top

# Temperature data

```{r temperature-raw, echo = TRUE}
temperatures <- read_csv(
  file = "~/github/webinar-msc-nhs/data/temp.csv",
  col_types = cols(date=col_date(format = "%d/%m/%Y"))
)

print(temperatures)
```

---

class: top

.sticker-float[![tsibble](resources/tsibble.svg)]

# Tidy temporal data structures

```{r}
htmltools::tags$iframe(src = "https://thesis.earo.me/3-3-sec-semantics.html", 
                       width = "100%", height = "500px")
```

---
class: top

.sticker-float[![tsibble](resources/tsibble.svg)]

# Tidy temperature data

Identify the index, key and measurement variables:

```{r temperature-qa, echo = TRUE, eval = FALSE}
temperatures <- as_tsibble(temperatures, 
           index = ???, key = ???) #<<
```

```{r}
temperatures
```

---
class: top

.sticker-float[![tsibble](resources/tsibble.svg)]

# Tidy temperature data

Identify the index, key and measurement variables:

```{r temperature-tsbl, echo = TRUE, eval = TRUE}
temperatures <- as_tsibble(temperatures, 
           index = date, key = NULL) #<<
```

```{r}
temperatures
```

---
class: top

.sticker-float[![tsibble](resources/tsibble.svg)]

# Tidy holiday data

Identify the index, key and measurement variables:

```{r holiday-qa, echo = TRUE, eval = FALSE}
holidays <- as_tsibble(holidays, 
           index = ???, key = ???) #<<
```

```{r}
holidays
```

---
class: top

.sticker-float[![tsibble](resources/tsibble.svg)]

# Tidy holiday data

Identify the index, key and measurement variables:

```{r holiday-tsbl, echo = TRUE, eval = TRUE}
holidays <- as_tsibble(holidays, 
           index = date, key = NULL) #<<
```

```{r}
holidays
```

---
class: top

.sticker-float[![tsibble](resources/tsibble.svg)]

# Tidy admissions data

Identify the index, key and measurement variables:

```{r admission-qa, echo = TRUE, eval = FALSE}
admissions <- as_tsibble(admissions, 
           index = ???, key = ???) #<<
```

```{r}
admissions
```

---
class: top

.sticker-float[![tsibble](resources/tsibble.svg)]

# Tidy admissions data

Identify the index, key and measurement variables:

```{r admission-bad, echo = TRUE, eval = TRUE, error = TRUE}
admissions <- as_tsibble(admissions, 
           index = arrival_time, key = c(gender, type_injury)) #<<
```

--

```{r admission-duplicates, echo = TRUE}
duplicates(admissions, index = arrival_time, key = c(gender, type_injury))
```

---
class: top

.sticker-float[![tsibble](resources/tsibble.svg)]

# Tidy admissions data

Identify the index, key and measurement variables:

```{r admission-tsbl, echo = TRUE, eval = TRUE, error = TRUE}
admissions <- admissions %>% 
  count(gender, type_injury, arrival_time, name = "arrivals") %>% #<<
  as_tsibble(index = arrival_time, key = c(gender, type_injury))
```

```{r}
admissions
```

---
class: top

.sticker-float[![tsibble](resources/tsibble.svg)]

# Bringing the data together

Can be a non-trivial task!

--

```{r tsbl-interval, echo = TRUE}
interval(admissions)
interval(holidays)
interval(temperatures)
```

---
class: top

.sticker-float[![tsibble](resources/tsibble.svg)]

# Choosing appropriate temporal intervals

What temporal granularity should `admissions` have?

--

A balancing act between:

* The temporal detail needed for the forecasts
* Enough detail/signal in the data for a good model

--

This choice has a big impact on the seasonal structures that will exist in the data!


---
class: top

.sticker-float[![tsibble](resources/tsibble.svg)]

# Choosing appropriate temporal intervals

Option A: Keep it as is (event level)

```{r temporal-a, echo = TRUE, fig.height=2.5}
admissions %>% 
  filter(yearmonth(arrival_time) == yearmonth("2010 Jan")) %>% 
  summarise(arrivals = sum(arrivals)) %>% #<<
  fill_gaps(arrivals = 0) %>% #<<
  autoplot()
#
#
```

---
class: top

.sticker-float[![tsibble](resources/tsibble.svg)]

# Choosing appropriate temporal intervals

Option B: 30 minutes

```{r temporal-b, echo = TRUE, fig.height=2.5}
admissions %>% 
  filter(yearmonth(arrival_time) == yearmonth("2010 Jan")) %>% 
  index_by(time = floor_date(arrival_time, "30 minutes")) %>% #<<
  summarise(arrivals = sum(arrivals)) %>% 
  fill_gaps(arrivals = 0) %>% 
  autoplot()
```

---
class: top

.sticker-float[![tsibble](resources/tsibble.svg)]

# Choosing appropriate temporal intervals

Option C: 1 hour

```{r temporal-c, echo = TRUE, fig.height=2.5}
admissions %>% 
  filter(yearmonth(arrival_time) == yearmonth("2010 Jan")) %>% 
  index_by(time = floor_date(arrival_time, "1 hour")) %>% #<<
  summarise(arrivals = sum(arrivals)) %>% 
  fill_gaps(arrivals = 0) %>% 
  autoplot()
```

---
class: top

.sticker-float[![tsibble](resources/tsibble.svg)]

# Choosing appropriate temporal intervals

Option D: 1 day

```{r temporal-d, echo = TRUE, fig.height=2.5}
admissions %>% 
  filter(year(arrival_time) == 2010) %>% 
  index_by(time = as.Date(floor_date(arrival_time, "1 day"))) %>% #<< 
  summarise(arrivals = sum(arrivals)) %>% 
  fill_gaps(arrivals = 0) %>% 
  autoplot()
```

---
class: top

.sticker-float[![tsibble](resources/tsibble.svg)]

# Choosing appropriate temporal intervals

Option E: 1 month

```{r temporal-e, echo = TRUE, fig.height=2.5}
admissions %>% 
  index_by(time = yearmonth(floor_date(arrival_time, "1 day"))) %>% #<<
  summarise(arrivals = sum(arrivals)) %>% 
  fill_gaps(arrivals = 0) %>%
  autoplot()
#
```

---
class: top

.sticker-float[![tsibble](resources/tsibble.svg)]

# Choosing appropriate temporal intervals

So what's best?

> * Has the temporal detail needed for the forecasts
> * Enough detail/signal in the data for a good model

--

.center[*It depends!*]

--

.center[
# 🙄
]

--

.center[For a multiple seasonality webinar, probably B or C. 

Let's go with hourly data (C) for as it has better signal.]

---

class: top

.sticker-float[![tsibble](resources/tsibble.svg)]

# Choosing appropriate temporal intervals

```{r admissions-hourly, echo = TRUE}
admissions <- admissions %>% 
  index_by(time = floor_date(arrival_time, "1 hour")) %>% #<<
  summarise(arrivals = sum(arrivals)) %>% 
  ungroup() %>% 
  fill_gaps(arrivals = 0)
```

--

```{r}
print(admissions)
```


---
class: middle
background-image: linear-gradient(to right, rgba(50, 50, 50, .5), rgba(50, 50, 50, .5)), url("resources/many_clocks.jpg")
background-size: cover

.box-12[
## Aside

Advanced forecasters may even choose multiple granularities!

<br>

Forecasts at many levels can be combined using **temporal reconciliation** to
get more accurate results. 

This process ensures the hourly forecasts add to the
daily forecasts, and so on.

<br>

`r fa("desktop")` ISF 2020 slides: [slides.mitchelloharawild.com/isf2020](https://slides.mitchelloharawild.com/isf2020)
`r fa("video")` Video recording: [youtu.be/6D7rNHZ5E-Q](https://youtu.be/6D7rNHZ5E-Q?t=1120)
]


---
class: top

.sticker-float[![tsibble](resources/tsibble.svg)]

# Bringing the data together

```{r tsbl-interval-update, echo = TRUE}
interval(admissions)
interval(holidays)
interval(temperatures)
```

--

It's okay for `holidays` to be daily as the holiday affects the entire day.

Ideally, `temperatures` would be hourly as it varies throughout the day.

---
class: top

.sticker-float[![tsibble](resources/tsibble.svg)]

# Bringing the data together

```{r, echo = TRUE}
admissions %>% 
  # Compute the common variable
  mutate(date = as.Date(time))
```

---
class: top

.sticker-float[![tsibble](resources/tsibble.svg)]

# Bringing the data together

```{r, echo = TRUE}
admissions %>% 
  # Compute the common variable
  mutate(date = as.Date(time)) %>% 
  # Join in holiday data (keeping only 'public_holiday' for simplicity)
  left_join(holidays %>% select(public_holiday), by = "date")
```

---
class: top

.sticker-float[![tsibble](resources/tsibble.svg)]

# Bringing the data together

```{r hospital, echo = TRUE}
hospital <- admissions %>% 
  # Compute the common variable
  mutate(date = as.Date(time)) %>% 
  # Join in holiday data (keeping only 'public_holiday' for simplicity)
  left_join(holidays %>% select(public_holiday), by = "date") %>% 
  # Join in temperature data (keeping only 'actual_temp' for simplicity)
  left_join(temperatures %>% select(actual_temp), by = "date")
```

```{r}
print(hospital)
```


---
class: inverse, center
background-image: linear-gradient(to right, rgba(50, 50, 50, .5), rgba(50, 50, 50, .5)), url("resources/graph.jpg")
background-size: cover

.title[Visualise]

<hr>

<br>

## Data exploration

---

# Visualising multiple seasonality

Plotting hourly data observed over six years can be tricky!

```{r admissions-plot, echo = TRUE, dev = "png", dpi = 150}
hospital %>% 
  autoplot(arrivals)
```

---

# Visualising multiple seasonality

One possibility is to view a portion of the data (but you might miss things!)

```{r hospital-dygraph, echo = TRUE}
library(dygraphs)
tsbox::ts_xts(hospital) %>% 
  dygraph() %>% 
  dyRangeSelector(dateWindow = c("2010-01-01", "2010-02-01"))
```

---

.sticker-float[![ggplot2](resources/feasts.svg)]

# Visualising multiple seasonality

Seeing the big picture may require aggregating the data.

```{r subseries-month, echo = TRUE, fig.height=2.5}
library(feasts)
hospital %>% 
  index_by(date = yearmonth(time)) %>% 
  summarise(arrivals = sum(arrivals)) %>% 
  gg_subseries(arrivals, period = "year")
```


---

.sticker-float[![ggplot2](resources/feasts.svg)]

# Visualising multiple seasonality

```{r subseries-week, echo = TRUE, dev = "png", dpi = 150}
hospital %>% 
  index_by(date = as.Date(time)) %>% 
  summarise(arrivals = sum(arrivals)) %>% 
  gg_subseries(arrivals, period = "week")
```

---

.sticker-float[![ggplot2](resources/feasts.svg)]

# Visualising multiple seasonality

```{r subseries-day, echo = TRUE, dev = "png", dpi = 150}
hospital %>% 
  gg_subseries(arrivals, period = "day")
```

---

.sticker-float[![ggplot2](resources/ggplot2.svg)]

# Visualising multiple seasonality

```{r ggplot-day, echo = TRUE, dev = "png", dpi = 150}
hospital %>% 
  ggplot(aes(x = hour(time), y = arrivals)) + 
  geom_jitter(alpha = 0.01) +
  geom_smooth() + 
  facet_grid(cols = vars(wday(time, label = TRUE, week_start = 1)))
```

---

.sticker-float[![ggplot2](resources/ggplot2.svg)]

# Visualising multiple seasonality

```{r cal-code, eval = FALSE, echo = TRUE}
library(sugrrants) #<<
p <- hospital %>% 
  mutate(date = as.Date(time)) %>% 
  filter(year(time) == 2012) %>% 
  frame_calendar(x = hour(time), y = arrivals, date = date) %>% #<<
  ggplot(aes(x = `.hour(time)`, y = .arrivals, group = date)) + 
  geom_line()
prettify(p)
```

---
.sticker-float[![ggplot2](resources/ggplot2.svg)]

# Visualising multiple seasonality

```{r cal-plot, ref.label="cal-code"}
```

---
class: inverse, center
background-image: linear-gradient(to right, rgba(50, 50, 50, .7), rgba(50, 50, 50, .7)), url("resources/books.jpg")
background-size: cover

.title[Making forecasts]

<hr>

<br>

## Data modelling

.left[
.footnote[
⚠️ The specific models/forecasts shown are in no way recommended or good.

They simply demonstrate how multiple seasonalities can be modelled.]
]

---

## Models for forecasting multiple seasonalities


<br>
<table class="table">
  <thead>
    <tr>
      <th>Model</th>
      <th>Flexible</th>
      <th>Speed</th>
      <th>Accuracy</th>
      <th>Decompose Components</th>
      <th>Evolving Terms</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <td>Regression<br>(MLR, GAM, <b>Prophet</b>)</td>
      <td class="success">Yes</td>
      <td class="success">Fast</td>
      <td class="warning">Okay</td>
      <td class="warning">Depends</td>
      <td class="danger">No</td>
    </tr>
    <tr>
      <td>State&nbspSpace<br>(DSHW, TBATS, <b>FASSTER</b>)</td>
      <td class="danger">No</td>
      <td class="danger">Slow</td>
      <td class="warning">Okay</td>
      <td class="success">Yes</td>
      <td class="success">Yes</td>
    </tr>
    <tr>
      <td>Decomposition<br>(<b>STL + ???</b>)</td>
      <td class="warning">Depends</td>
      <td class="warning">Moderate</td>
      <td class="warning">Okay</td>
      <td class="success">Yes</td>
      <td class="warning">Depends</td>
    </tr>
  </tbody>
</table>

.footnote[(This table is a *massive* generalisation, everything depends on data and model complexity.)]

---

## Regression modelling

<br>

<table class="table">
  <thead>
    <tr>
      <th>Model</th>
      <th>Flexible</th>
      <th>Speed</th>
      <th>Accuracy</th>
      <th>Decompose Components</th>
      <th>Evolving Terms</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <td>Regression<br>(MLR, GAM, <b>Prophet</b>)</td>
      <td class="success">Yes</td>
      <td class="success">Fast</td>
      <td class="warning">Okay</td>
      <td class="warning">Depends</td>
      <td class="danger">No</td>
    </tr>
  </tbody>
</table>

Uses time series features / regressors with known future values to produce forecasts.

✅ Generally fast, as model estimation is time independent.

❌ Model behaviour does not update over time for changing dynamics.

Highly recommended read on GAM forecasting: https://petolau.github.io/Analyzing-double-seasonal-time-series-with-GAM-in-R/
---

## Prophet

Facebook's forecasting model based on a GAM which uses regressors for growth, seasonality, and holidays.

$$
  y(t) = g(t) + s_1(t) + s_2(t) + h(t) + \varepsilon_t
$$

where $g(t)$ is the growth function, $s(t)$ are fourier terms, and $h(t)$ are holiday regressors.

Available directly in the [`{prophet}`](https://facebook.github.io/prophet/) package, or a fable interface from [`{fable.prophet}`](http://pkg.mitchelloharawild.com/fable.prophet/).

---

## Specifying a prophet model

Using default settings

```{r}
library(fable.prophet)
hospital %>% 
  model(prophet(arrivals)) %>% 
  forecast(h = "2 weeks") %>% 
  autoplot(tail(hospital, 24*7*4))
```

---

## Specifying a prophet model

Customising the `season()` options.

```{r}
library(fable.prophet)
hospital %>% 
  model(prophet(arrivals ~ season("day", 7, type = "mult") + season("week", 4))) %>% 
  forecast(h = "2 weeks") %>% 
  autoplot(tail(hospital, 24*7*4))
```

---

## State space modelling

<br>

<table class="table">
  <thead>
    <tr>
      <th>Model</th>
      <th>Flexible</th>
      <th>Speed</th>
      <th>Accuracy</th>
      <th>Decompose Components</th>
      <th>Evolving Terms</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <td>State&nbspSpace<br>(DSHW, TBATS, <b>FASSTER</b>)</td>
      <td class="danger">No</td>
      <td class="danger">Slow</td>
      <td class="warning">Okay</td>
      <td class="success">Yes</td>
      <td class="success">Yes</td>
    </tr>
  </tbody>
</table>

Iterates 'state equations' over time to produce forecasts.

✅ Model behaviour updates for changes in the data.

❌ Relatively slow, as model estimation is done sequentially over time.

---

## FASSTER

A model which provides flexible regression-style model specification with evolving terms of state space models via a DLM.

\begin{align*}
y_t &= F_t\theta_t + v_t, & v_t &\sim \mathcal{N}(0,V)\\
\theta_t &= G\theta_{t-1} + w_t, & w_t&\sim \mathcal{N}(0,W)
\end{align*}

The underlying states are given by $\theta$, which can be thought of as components linearly combined by $F_t$ to produce the response $y_t$. The $G$ matrix defines the behaviour of each state (seasonality, trend, regression, ...).

Available with direct fable compatibility in the [`{fasster}`](http://fasster.tidyverts.org/) package.

---

## Specifying seasonality with FASSTER

FASSTER allows all aforementioned seasonal terms to be specified.

```{r fasster-seasonality}
tibble::tribble(
  ~ `Seasonal term`, ~ Special,
  "Seasonal dummies", "y ~ season(p)",
  "Lag terms", "y ~ lag(y, p)",
  "Fourier terms", "y ~ fourier(p, K)",
  "Regressors", "y ~ x"
) %>% 
  gt::gt()
```

---

## Specifying several seasonalities with FASSTER

```{r fasster-simple}
library(fasster)
ped_train <- pedestrian %>% 
  filter(Sensor == "Southern Cross Station",
         yearmonth(Date) == yearmonth("2015 Apr"))
ped_train %>% 
  model(fasster(Count ~ trend(1) + fourier("week", 3) + (wday(Date_Time) %in% c(1,7)) %S% fourier("day", 10))) %>% 
  forecast(h = "2 weeks") %>% 
  autoplot(ped_train)
```

---

## Decomposition modelling

<br>

<table class="table">
  <thead>
    <tr>
      <th>Model</th>
      <th>Flexible</th>
      <th>Speed</th>
      <th>Accuracy</th>
      <th>Decompose Components</th>
      <th>Evolving Terms</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <td>Decomposition<br>(<b>STL + ???</b>)</td>
      <td class="warning">Depends</td>
      <td class="warning">Moderate</td>
      <td class="warning">Okay</td>
      <td class="success">Yes</td>
      <td class="warning">Depends</td>
    </tr>
  </tbody>
</table>

Decomposes time series into components, allowing seasonality to be forecasted
separately with simpler methods.

✅ Allows complex time series to be forecasted with simpler models.

❌ Highly dependent on the ability to decompose the time series.

---

## STL + ???

Uses repeated STL decompositions to decompose multiple seasonalities into separate time series. Then other models (`???`) can be used to forecast the simpler components.

$$y_{t} = S_{\text{year},t} + S_{\text{week},t} + S_{\text{day},t} + T_{t} + R_t$$

Available with direct fable compatibility in the [`{feasts}`](http://feasts.tidyverts.org/) package.

---

## Decomposing a time series

Using an `STL` decomposition with default settings:

```{r dcmp-stl, echo = TRUE, dev = "png", dpi = 150}
library(feasts)
hospital_dcmp <- hospital %>% 
  model(STL(arrivals)) 
components(hospital_dcmp) %>% autoplot()
```

---

## Decomposing a time series

Using an `STL` decomposition with default settings:

```{r dcmp-stl-plot, echo = TRUE}
components(hospital_dcmp) %>% 
  tail(24*7*5) %>% 
  autoplot()
```

---

## Decomposing a time series

Refining the loess windows for the sub-daily data frequency:

```{r dcmp-custom, echo = TRUE, dev = "png", dpi = 150}
hospital_dcmp <- hospital %>% 
  model(STL(arrivals ~ trend(window = 24*7*5) + season("year", window = Inf) +
              season("week", window = 24*7*5) + season("day", window = 24*7)))
components(hospital_dcmp) %>% autoplot()
```

---

## Decomposing a time series

Refining the loess windows for the sub-daily data frequency:

```{r dcmp-custom-plot, echo = TRUE}
components(hospital_dcmp) %>% 
  tail(24*7*5) %>% 
  autoplot()
```

---

## Modelling decomposed components

As seasonality changes slowly, it is easily modelled with naive lags.

```{r season-week-fc, echo = TRUE}
components(hospital_dcmp)[-1] %>% 
  model(SNAIVE(season_week ~ lag("week"))) %>% 
  forecast(h = "2 weeks") %>% 
  autoplot(tail(components(hospital_dcmp)[-1], 24*7*4))
```

---

## Modelling decomposed components

As seasonality changes slowly, it is easily modelled with naive lags.

```{r season-day-fc, echo = TRUE}
components(hospital_dcmp)[-1] %>% 
  model(SNAIVE(season_day ~ lag("day"))) %>% 
  forecast(h = "2 weeks") %>% 
  autoplot(tail(components(hospital_dcmp)[-1], 24*7*4))
```

---

## Modelling decomposed components

The seasonally adjusted data (trend + remainder) is typically forecasted together.

```{r seasadjust-arima, echo = TRUE}
components(hospital_dcmp)[-1] %>% 
  model(ARIMA(season_adjust ~ 0 + pdq(3,0,3) + PDQ(1,0,0))) %>% 
  forecast(h = "3 days") %>% 
  autoplot(tail(components(hospital_dcmp)[-1], 24*4))
```

---

## Combining individual forecasts

Use `decomposition_model()` to specify all component models:

```{r dcmp-model, echo = TRUE}
dcmp_spec <- decomposition_model(
  dcmp = STL(arrivals ~ trend(window = 24*7*5) + season("year", window = Inf) +
              season("week", window = 24*7*5) + season("day", window = 24*7)),
  ARIMA(season_adjust ~ 0 + pdq(3,0,3) + PDQ(1,0,0)),
  SNAIVE(season_week ~ lag("week")),
  SNAIVE(season_day ~ lag("day"))
)
```

--

Notice that I haven't provided a model for `season_year`. A `SNAIVE()` model will
be used by default for seasonal components.

---

## Combining individual forecasts

Then use it as a regular model, the recombination is done automatically.

```{r dcmp-model-fc, echo = TRUE}
hospital %>% 
  model(dcmp_spec) %>% 
  forecast(h = "2 weeks") %>% 
  autoplot(tail(hospital, 24*7*4))
```


---
class: inverse, top
background-image: linear-gradient(to right, rgba(150, 150, 150, .1), rgba(150, 150, 150, .4)), url("resources/hourglass.jpg")
background-size: cover

.sticker-float[![fable](resources/fable.svg)]

.title[Thanks! `r fa("comments", fill = "white")`]

.larger[
`r fa("globe", fill = "white")` Learn more: [fable.tidyverts.org](https://fable.tidyverts.org/)

`r fa("chart-line", fill = "white")` Keep updated: [tidyverts.org](http://www.tidyverts.org)

`r fa("desktop", fill = "white")` Review slides: [slides.mitchelloharawild.com/nhs2020](https://slides.mitchelloharawild.com/nhs2020)

`r fa("twitter", fill = "white")` Say hello: [@mitchoharawild](twitter.com/mitchoharawild/)

<br>

.bottom[This work is licensed as `r fa("creative-commons", fill="white")` BY-NC 4.0.]
]
